---
title: "Reproducible Research: Peer Assessment 1"
output: 
  html_document:
    keep_md: true
---


## Loading and preprocessing the data

The data will be unzipped and loaded into a table. The only preprocessing we will apply is to convert the dates to Date type.

```{r, cache=TRUE}
data <- read.csv(unz("activity.zip", "activity.csv"))
data$date <- as.Date(data$date)
summary(data)
```

## What is mean total number of steps taken per day?

In order to obtain the total steps per day we will use the aggregate function. This function will group the number of steps by date and then sum them. Those rows with a number of steps = 0 will be deleted, as it means no activity (we are monitoring activity).  
 

```{r}
steps_per_day <- aggregate(data$steps, by=list(date=data$date), FUN=sum, na.rm = TRUE)
# Remove those with 0 steps (no activity that day)
positive_steps_per_day <- subset(steps_per_day, x!=0)
```

We then proceed to plot the data. Mean and median are calculated, and mean is also plotted.    

```{r}
library(ggplot2)
ggplot(data=positive_steps_per_day, aes(positive_steps_per_day$x)) + 
    geom_histogram(bins = 25)+
    labs(x="Steps per day (>0)", y="Count")+
    geom_vline(aes(xintercept=mean(x)),color="red", linetype="dashed", size=1)
```
```{r, echo=FALSE}
m1 <- mean(positive_steps_per_day$x)
md1 <- median(positive_steps_per_day$x)
```

The mean is `r m1` and the median is `r md1`.

## What is the average daily activity pattern?

In order to answer this queston we now have to aggregate per interval. Then we can show the time series usign a regular plot.  
We define a function with the plotting code, as it will be reused later.  

```{r}
plot_timeline <- function(data, title="", ylim=250){
    avg_steps_per_interval <- aggregate(data$steps, by=list(interval=data$interval), FUN=mean, na.rm = TRUE)
    
    max_interval_index <- which(avg_steps_per_interval$x == max(avg_steps_per_interval$x))
    max_interval <- avg_steps_per_interval[max_interval_index,]
    
    ggplot(avg_steps_per_interval, aes(interval,x)) + 
        geom_line() +
        labs(x="5' Interval", y="Avg. Num. Steps", title=title)+
        geom_vline(aes(xintercept=max_interval$interval),color="red", linetype="dashed", size=1)+
        ylim(0,ylim)
}

plot_timeline(data)
```

## Imputing missing values

In order to eliminate the missing values, we will exchange them by the averages over that 5' time period (*avg_steps_per_interval*).    

We first calculate the number of NAs:
```{r}
number_of_nas <- sum(is.na(data$steps))
```
which is `r number_of_nas` (the same value we have seen in the initial summary).  

Next, we calculate the *avg_steps_per_interval*, and change the values of the records containing NA.  

```{r}
avg_steps_per_interval <- aggregate(data$steps, by=list(interval=data$interval), FUN=mean, na.rm = TRUE)

imputed_data <- data
for (i in 1:dim(imputed_data)[1]){
    if (is.na(imputed_data[i,"steps"])){
        interval <- imputed_data[i,"interval"] # 5*(i-1) indeed, or 5*i if we start at 0
        interval_index <- which (avg_steps_per_interval$interval == interval)
        imputed_data[i,"steps"] <- avg_steps_per_interval$x[interval_index]
    }
}
```

Then we can recalculate the values we will plot in the histogram with the same method explained in answer 1. The new mean and median are calculated afterwards.  

```{r}
# redo calulations
imp_steps_per_day <- aggregate(imputed_data$steps, 
                               by=list(date=imputed_data$date), 
                               FUN=sum, na.rm = TRUE)


# Remove those with 0 steps (no activity that day)
imp_positive_steps_per_day <- subset(imp_steps_per_day, x!=0)

ggplot(data=imp_positive_steps_per_day, aes(imp_positive_steps_per_day$x)) + 
    geom_histogram(bins = 25) +
    labs(x="Steps per day (>0)", y="Count") +
    geom_vline(aes(xintercept=mean(x)),color="red", 
               linetype="dashed", size=1)
```

```{r, echo=FALSE}
m2 = mean(imp_positive_steps_per_day$x)
md2 = median(imp_positive_steps_per_day$x)
```

The value for the mean is now `r m2` instead of  `r m1`  and the value for the median is `r md2` instead of  `r md1`. Surprisingly, the values have not changed too much.

## Are there differences in activity patterns between weekdays and weekends?

In order to answer this last question, we first define a function to tell us if a given date is worweek day or weekend. 

```{r}
day_type <- function(date){
     if (weekdays(date) %in% c("Saturday","Sunday")){
         return("weekend")
     }
    else{
        return("weekday")
    };
}
```

We can apply this function to the table in order to obtain the new column.

```{r}
imputed_data$day.type <- sapply(imputed_data$date, day_type, simplify=TRUE)
```

This allows us to easily extract the data for each of the day types using the *subset* operation.

```{r}
weekday_data = subset(imputed_data, day.type == "weekday")
weekend_data = subset(imputed_data, day.type == "weekend")
```

And we just need to plot the results (adjusting the y scale in order to make the comparison easier):

```{r}
plot1 <- plot_timeline(weekday_data, "Workday", 250)
plot2 <- plot_timeline(weekend_data, "Weekend", 250)
library(gridExtra)
g = grid.arrange(plot1, plot2, ncol=2,  newpage = TRUE)
```

The plots show more activity during the weekend, but the activity peaks are placed in the same time window.
